---
title: "Getting Started with TweedSpot"
author: "Nalin Arora"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Getting Started with TweedSpot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

TweedSpot detects spatially variable genes (SVGs) in spatial transcriptomics
data using Tweedie generalized additive models (GAMs) with a spatial smooth.
The package currently focuses on an agnostic SVG workflow built around the
`SpatialExperiment` container.

This vignette covers a standard workflow:

1. load a `SpatialExperiment`
2. prefilter low-information genes
3. fit TweedSpot
4. inspect gene-level results

# Load packages

```{r libraries}
library(TweedSpot)
library(SpatialExperiment)
library(SummarizedExperiment)
library(STexampleData)
```

# Load example data

We use the `Visium_humanDLPFC()` example dataset from `STexampleData`. To keep
the vignette lightweight, we restrict the analysis to in-tissue spots and a
small subset of genes.

```{r load-data}
spe <- Visium_humanDLPFC()
spe <- spe[, colData(spe)$in_tissue == 1]
spe <- spe[seq_len(min(100, nrow(spe))), ]

spe
```

# Prefilter genes

`filter_genes()` removes genes with limited information before model fitting.
In addition to minimum total counts and percent of nonzero spots, it can also
filter on the number of nonzero spots, mean counts, variance, and optionally
exclude mitochondrial or ribosomal genes.

```{r filter-genes}
spe <- filter_genes(
  input = spe,
  assay_name = "counts",
  filter_genes_ncounts = 5,
  filter_genes_pcspots = 5,
  filter_genes_nspots = 3,
  filter_genes_mean = 0.05
)

spe
```

# Fit TweedSpot

We now fit TweedSpot on the filtered dataset. Here we adjust for `sample_name`
from `colData(spe)` and use a smaller `smooth_k` to keep the example fast.

```{r run-tweedspot}
spe_res <- tweedspot(
  input = spe,
  assay_name = "counts",
  covariates = ~ sample_name,
  two_part = FALSE,
  family = "tw",
  fit_method = "REML",
  use_bam = TRUE,
  bam_discrete = TRUE,
  smooth_k = 10
)
```

# Inspect results

TweedSpot writes per-gene outputs into `rowData(input)`. The main result
columns are:

- `tweedspot_stat`: spatial smooth test statistic
- `tweedspot_pval`: raw p-value
- `tweedspot_padj`: adjusted p-value
- `tweedspot_edf`: effective degrees of freedom of the spatial smooth
- `tweedspot_dev_expl`: deviance explained by the fitted GAM

```{r inspect-results}
res <- as.data.frame(rowData(spe_res))

head(res[, c(
  "tweedspot_stat",
  "tweedspot_pval",
  "tweedspot_padj",
  "tweedspot_edf",
  "tweedspot_dev_expl"
)])
```

We can also rank genes by adjusted p-value:

```{r top-genes}
top_idx <- order(res$tweedspot_padj, decreasing = FALSE, na.last = NA)
head(res[top_idx, c(
  "tweedspot_pval",
  "tweedspot_padj",
  "tweedspot_edf",
  "tweedspot_dev_expl"
)])
```

# Notes

- TweedSpot currently supports the agnostic SVG setting.
- Covariates are specified from `colData(input)` using a one-sided formula such
  as `~ batch + age + sex`.
- If `colData(input)$sizeFactor` is present, TweedSpot uses it as the library
  size adjustment. Otherwise, it falls back to total-count scaling.

# Session information

```{r session-info}
sessionInfo()
```
