---
title: "TweedSpot"
author: "Nalin Arora"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Getting Started with TweedSpot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

TweedSpot detects spatially variable genes (SVGs) in spatial transcriptomics
data using Tweedie generalized additive models (GAMs) with a spatial smooth.
The package currently focuses on an agnostic SVG workflow built around the
`SpatialExperiment` container.

This vignette covers a standard workflow:

1. load a `SpatialExperiment`
2. prefilter low-information genes
3. fit TweedSpot
4. extract and rank gene-level results
5. visualize observed and fitted spatial patterns

# Load packages

```{r libraries}
library(TweedSpot)
library(SpatialExperiment)
library(SummarizedExperiment)
library(STexampleData)
library(BiocParallel)
```

# Load example data

We use the `Visium_humanDLPFC()` example dataset from `STexampleData`. To keep
the vignette lightweight, we restrict the analysis to in-tissue spots and a
small subset of genes.

```{r load-data}
spe <- Visium_humanDLPFC()
spe <- spe[, colData(spe)$in_tissue == 1]
spe <- spe[seq_len(min(250, nrow(spe))), ]
spe
```

# Prefilter genes

`filter_genes()` removes genes with limited information before model fitting.
In addition to minimum total counts and percent of nonzero spots, it can also
filter on the number of nonzero spots, mean counts, variance, and optionally
exclude mitochondrial or ribosomal genes. By default it also drops spots with
zero total counts after gene filtering.

```{r filter-genes}
spe <- TweedSpot::filter_genes(
  input = spe,
  assay_name = "counts",
  filter_genes_ncounts = 5,
  filter_genes_pcspots = 5,
  filter_genes_nspots = 3,
  filter_genes_mean = 0.05,
  drop_zero_spots = TRUE,
  report = TRUE
)

spe
```

# Fit TweedSpot

We now fit TweedSpot on the filtered dataset. For this example object,
`sample_id` is constant across all retained spots, so we do not include it as a
covariate. We also use a smaller `smooth_k` to keep the example fast.

```{r run-tweedspot}


bp <- MulticoreParam(workers = 4)

spe_res <- tweedspot(
  input = spe,
  assay_name = "counts",
  two_part = FALSE,
  family = "tw",
  fit_method = "REML",
  use_bam = TRUE,
  bam_discrete = TRUE,
  smooth_k = 10,
  BPPARAM = bp,
  verbose = TRUE
)
```

# Extract and rank results

TweedSpot writes per-gene outputs into `rowData(input)`. The main result
columns are:

- `tweedspot_stat`: spatial smooth test statistic
- `tweedspot_pval`: raw p-value
- `tweedspot_padj`: adjusted p-value
- `tweedspot_edf`: effective degrees of freedom of the spatial smooth
- `tweedspot_dev_expl`: deviance explained by the fitted GAM

`get_tweedspot_results()` returns a clean `data.frame` representation of these
results. By default it sorts by adjusted p-value, breaking ties by
`tweedspot_stat`.

```{r inspect-results}
res <- get_tweedspot_results(spe_res)

head(res)
```

`top_spatial_genes()` provides a convenience wrapper for retrieving the top
hits by adjusted p-value, raw p-value, statistic, or deviance explained.

```{r top-genes}
top_spatial_genes(spe_res, n = 10, rank_by = "padj")
```

# Visualize observed and fitted spatial patterns

The package also provides plotting helpers for the raw observed expression
pattern and the fitted spatial smooth for a single gene.

```{r plot-spatial, fig.width=10, fig.height=4}
top_gene <- res$gene_name[1]

plot_spatial_gene(spe_res, gene = top_gene)
plot_spatial_fit(spe_res, gene = top_gene, family = "tw", smooth_k = 10)
```

# Notes

- TweedSpot currently supports the agnostic SVG setting.
- Covariates are specified from `colData(input)` using a one-sided formula such
  as `~ batch + age + sex`.
- If `colData(input)$sizeFactor` is present, TweedSpot uses it as the library
  size adjustment. Otherwise, it falls back to total-count scaling.
- `filter_genes()` can optionally report how many genes and spots were retained
  and how many were removed by each criterion.
- `tweedspot()` can print lightweight progress messages through
  `verbose = TRUE`.

# Session information

```{r session-info}
sessionInfo()
```
